\section{Introduction}
\label{sec:introMain}

Developers spend a substantial portion of their time understanding source code---that is, interpreting its meaning to work effectively with it---with % MK: you must use em-dashes to offset this phrase and not commas, because the offset phrase itself contains a comma
prior studies estimating that between 58\% and 70\% of their time is devoted to code comprehension~\cite{percent1,scalabrino2017automatically}. As a result, there is strong motivation among researchers to improve this process through scientific study. This interest is reflected in venues such as the International Conference on Program Comprehension (ICPC), which recently held its 34th edition in April 2026, in Rio de Janeiro, Brazil.

Whether the goal is to understand how source code can be written more comprehensibly or to identify which contextual factors and individual characteristics influence code comprehension, there is a clear need for reliable ways to measure it.

However, the construct of code comprehension lacks a widely accepted and practically operational definition for empirical study design, let alone a standardized measurement method or clear consensus on which features should be included~\cite{Wyrich:ACS23}. As a result, researchers often define comprehensibility in terms of the measurements they employ.

Furthermore, because comprehensibility is a cognitive process occurring in developers’ minds, it cannot be measured directly. Consequently, studies in the area of code comprehensibility mostly rely on indirect proxies\todo{do you want to italicize ``proxies'' as a techincal term? I would do so, and define a proxy explicitly as a pair of a measurement and a task.}: observable \textit{measurements} collected from humans performing specific comprehension \textit{tasks}. Examples of such tasks include summarizing code~\cite{S4, S11, S24, S35, 33-7, 33-8, S38, S42, S49} and fixing bugs~\cite{S38, S55, S64, S87}. Common measurements include subjective ratings of code understandability~\cite{S4, S34, 33-7, S42, S49, 33-20}, the time required to answer questions about the code~\cite{Scalabrino:TSE19, S11, S39, S48, S50, S57, S59, S60, 33-24}, and the correctness of those answers~\cite{Scalabrino:TSE19, S11, NovicesVSprof1:BIBE2016, S48, S50, S57, S59, S60, 33-24}.

These proxies are intended to capture the difficulty of comprehension (i.e., ``code comprehensibility''). For example, longer response times are typically interpreted as indicating that a code snippet is harder to understand. Comprehension proxies have been used not only to investigate how different factors influence comprehensibility~\cite{S4,S49,Peitek:TSE18,Siegmund:ICSE14,Stapleton:ICPC2020,Wyrich:ACS23}, but also to develop predictive models that help developers estimate and manage comprehension difficulty during software evolution~\cite{Scalabrino:TSE19,Raymond:TSE10,Lavazza:ESE2023, De:2025}. Despite their widespread use, however, the reliability of these proxies has remained largely unexamined.

In our recent study\todo{refer to this study as ``the first part of this work'' or similar, to make it clear that it's part of your thesis and not a separate line of work} (see Section~\ref{ase}), we aimed to address two major problems in code comprehensibility research: the lack of a practical conceptualization of comprehensibility and the absence of a systematic evaluation of commonly used proxies in human studies. We conceptualized code comprehensibility based on expert agreement about comprehension difficulty, enabling the construction of ground-truth data. We conducted an expert study using the Delphi protocol to establish consensus-based rankings of code snippets. We then ran a study with student participants to collect 14 comprehension proxies commonly used in prior work. Finally, we performed a correlation analysis comparing these proxies against expert-consensus judgments, showing that semantic-based proxies—especially the time required to correctly determine a program’s output for a given input—align more closely with expert assessments, while syntactic proxies (i.e., response time or accuracy on syntactic questions about the code) tend to exhibit the weakest—and in some cases, negative—correlations with comprehensibility.

These findings have important implications: they provide a practical and more reliable way to approximate code comprehensibility in human studies, facilitating more rigorous research in this area.

\todo{There is a logical break in the argument here. Maybe add subsections?}

However, conducting human studies remains expensive and time-consuming. Such studies must be carefully designed to isolate the relationship between code comprehensibility and other variables, and they can take months to prepare. Every aspect of the process requires careful attention, from constructing experimental materials and defining participant criteria to designing protocols and analyzing results. Running these studies is also costly, requiring participant recruitment, compensation, and often supervised sessions.

As a result, it is not feasible to rely on human studies whenever we need to measure code comprehensibility—for example, to compare which version of an implementation is more understandable. More importantly, the increasing need for automation in software engineering makes it essential to develop reliable automatic proxies for code comprehensibility (e.g., for evaluating and improving LLM-generated code in terms of comprehensibility). Such proxies would enable tools to identify hard-to-understand code and automatically suggest or perform refactorings.

Prior work has proposed many syntactic (i.e., non-semantic\todo{I would explain what ``syntactic'' means in plain(er) English rather than by referring to semantics here, since anyone who knows what ``semantics'' means presumably also knows what ``syntax'' means.}) metrics to approximate code comprehensibility without the need for human studies~\cite{ajami2019syntax, chidamber1994metrics, curtis2006measuring, henderson1995object, jbara2017programmers, nunez2017source, sneed1995understanding,  zuse1993criteria}. However, recent studies~\cite{feigenspan2011exploring, peitek2021program, scalabrino2019automatically} show that such metrics often correlate weakly, if at all, with developers’ brain activity and behavior while understanding code. Our own findings in~\cref{ase} further reinforce this limitation, showing that syntactic-based proxies have the lowest predictive power among commonly used measures. This raises doubts about the feasibility of purely syntactic automatic proxies for code comprehensibility.

This naturally leads to the question: can we incorporate semantic information into automatic proxies for code comprehensibility?

To explore this direction, recent work in our group~\cite{FeldmanKC2023}\todo{this study you should refer to as ``prior work''; you do not need to mention that it was in our group} investigated the hypothesis that code that is easier for verification tools to analyze is also easier for humans to understand. In that work, verifiability was defined as the effort required to establish safety properties (e.g., the absence of null pointer dereferences or out-of-bounds accesses) using verification tools (e.g., checkers implemented in the Checker Framework~\cite{checkerframework-manual}\todo{probably better to cite~\cite{PapiACPE2008}}), and was operationalized via the presence or absence of false positive warnings. A meta-analysis was conducted over prior comprehensibility studies, aggregating results across different tools, warning types, and experimental settings, and revealing associations between verification-related signals and measures of human understanding.

However, this approach has an inherent limitation: because it relies on heterogeneous prior studies, it does not allow us to isolate the relationship between code comprehensibility and code verifiability.
\todo{Mention that the results also had wide confidence intervals, so there is some question about whether the effect actually exists or whether it is just an artifact in the data. You can also cite Nadeeshan's EASE paper to help make this point~\cite{DeSilvaKC2026}.}

To address this limitation, I\todo{always ``we'', even when you are sole author. A quirk of academic English, specifically.} propose a controlled experimental study (see Section~\ref{sec:nextpaper}\todo{here and elsewhere, use \<cref> instead of \<ref> when refering to sections, so that you don't need to explicitly write ``Section'' in the text.}) that systematically manipulates verifiability while constraining other sources of variation. We fix the verification configuration by selecting a small set of tools, construct semantically equivalent code variants that differ in verifiability while minimizing structural differences through carefully designed transformations, and reduce variability through randomized assignment and appropriate statistical modeling. In this controlled setting, we hold constant the verification configuration, code semantics and structure, dataset characteristics, comprehension tasks, and participant population; we standardize measurements, control for order effects, and account for residual variability statistically. This design enables a more precise examination of the relationship between verifiability and comprehensibility.

If a strong relationship is established, verifiability could serve as a practical automatic proxy, reducing reliance on costly human studies and enabling scalable tools for improving code quality. Even a moderate relationship would support incorporating verifiability as a component in multi-factor proxies.\todo{Add a sentence describing the implication if you find a negative result: you can refute our FSE 23 paper. This implication is why it's important to point out earlier that the result from that paper is rather tenuous.}

In~\cref{sec:futurework}, I outline several promising future directions, including evaluating whether LLM agents can reliably simulate human comprehension behavior and conducting a systematic reassessment of prior studies in light of our findings on proxy reliability.

The rest of this proposal is structured as follows. Section~\ref{sec:relatedworkMain} provides general background and related work. Section~\ref{ase} presents the paper we submitted to ASE 2026, which examines the reliability of comprehension proxies in human studies. Section~\ref{sec:nextpaper} describes my proposed next paper\todo{Be a bit less explicit: ``the follow up study that we plan to carry out next'' or something like that rather than referring to the next paper explicitly. You should also say a little bit about it here to connect it in the reader's mind to the experiment you were describing just above.}. Section~\ref{sec:futurework} outlines several directions for future work beyond the next paper. Finally, Section~\ref{sec:conclusion} provides an overview of the proposal.\todo{You don't need to signpost the conclusion.}\todo{I would prefer this as a list of contributions, but if you prefer this paragraph format it's okay too.}
